\huge{Male-biased sexual selection, but not sexual dichromatism, predicts speciation

### 2019-07-12 Manuscript_Publication.Rmd
changes to figures

---
title: \huge{Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds}
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§}
  and Iliana Medina Guzmán\textsuperscript{§}
- ~
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  pdf_document: default
  word_document: default
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape} 
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} 
  \usepackage[hang, flushmargin]{footmisc} 
  \usepackage{hyperref}\hypersetup{colorlinks=true, hyperfootnotes=false, linkcolor=blue, filecolor=blue, urlcolor=blue}
  \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: 'Sexual selection can shape phylogenetic diversity by altering speciation
  or extinction rates. The predicted effects on diversity are often contradictory;
  for example, sexual selection might promote speciation by creating reproductive
  isolation, or hinder it by selecting for traits that facilitate gene flow. Similarly,
  sexual selection might precipitate extinction by favouring competitive phenotypes
  with poor survival or stave it off by helping populations shed deleterious mutations.
  Here, we investigate the relationship between sexual selection and rates of speciation
  and extinction in passerine birds. We use two measures of sexual selection --- sexual
  dichromatism (*n* = 5,812 species) and a multivariate measure of male-biased sexual
  selection (sexual dimorphism, polygyny and paternal care; *n* = 2,465) --- as well
  as three distinct measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$),
  and one measure of extinction ($\mu_{BAMM}$). We also tested whether the effects
  of sexual selection is contingent on ecological variables such as spatiotemporal
  temperature variation, primary productivity, and range size. We find that birds
  with smaller ranges have higher speciation rates, although the direction of causality
  is unclear. Moreover, our results show that male-biased sexual selection, but not
  sexual dichromatism, is an agent of adaptation and speciation. This finding suggests
  that sexual selection is an evolutionary force shaping patterns of diversity amongst
  songbirds, which has broad implications for ecology and conservation biology. \vspace{5mm}  \par\noindent \textbf{Keywords:} Speciation rate, Comparative analysis, Plumage dichromatism, Range size, Extinction rate.'
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

#Introduction

Selection pressures are the main engine of speciation. As populations diverge they accumulate genetic and phenotypic changes that ultimately erect barriers against gene flow [@Presgraves_2010; *but see* @Shaw_2011]. In this view, populations will speciate and adapt more readily to new and different environments if selection is strong and/or persistent. Hence, the rate at which speciation occurs is likely dependent on both the population's response to selection (adaptation) and the potential for divergent phenotypic traits to evolve [*e.g.*, morphology or colour: @Hugall_2012; @Rabosky_morph_2013]. Experiments suggest that sexual selection can influence the evolution of a surprisingly diverse set of traits [@Cally_in_review], and may therefore strongly influence extant patterns of phylogenetic diversity.

Sexual selection results from competition for matings, fertilisations, or associated resources, and is typically stronger in males than females [@Fromhage_2016; @Janicke_2016; @Kokko_Jennions2008]. This evolutionary force has been proposed to facilitate speciation through the evolution of divergent signals associated with improved mating success [@Lande_1981; @Lande_1982]. For instance, in antbirds (Thamnophilidae), genera with complex songs and striking dichromatism are more species-rich [@Seddon_2008]. Under this model of speciation, divergent mating signals in one sex (usually males) co-evolve with divergent preferences for those signals in the opposite sex, leading to behavioural reproductive isolation [@Safran_2013]. Alongside facilitating the evolution of divergent mating signals, sexual selection can increase the extent populations diverge by enlarging the available phenotypic space for mating signals. These predictions emerge as sexual selection may increase genetic diversity as a result of trade-offs between naturally- and sexually-selected traits or trade-offs under *intra*-locus sexual conflict between male and female trait expressions [@Lorch_2003; @Bonduriansky_2011; @Radwan_2016]. Additionally, sexual selection can have both positive and negative consequences for adaptation [*e.g.*, @Kokko_Brooks_2003; @Whitlock_Agrawal_2009; @Holman_Kokko_2013]. For example, sexual selection can elevate average fitness by helping to purge deleterious mutations [@Agrawal_2001; @Siller_2001] and fix beneficial ones [@Whitlock_2000], which should mitigate extinction risk and facilitate adaptation to environmental change. However, sexual selection might also promote extinction by selecting for male traits that have detrimental pleiotropic effects on female traits [@Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013; @Pischedda_2006], or selecting for male phenotypes with improved mating success but which harm population productivity, such as under-investing in parental care [@RN277].

Importantly, the effects of sexual selection on adaptation and speciation may depend on environmental conditions. Both theoretical and empirical studies suggest that under stressful/changing environments, sexual selection may have greater fitness benefits than under benign environments. Stressful environments strengthen the positive correlation between male mating success and female fitness, reducing the burden of negative pleiotropic effects [@Berger_2014; @Connallon_Hall_2016; @Long_2012; @Martinossi-Allibert_Savković_2017]. Conversely, benign or stable environments ensure consistent selection, preferentially eroding genetic variation at sexually concordant loci (*i.e.* loci where the optimum genotype is equivalent for both sexes). In these stable environments, genetic variation remains disproportionately at sexually antagonistic loci; causing displacement of male and female traits from their optimum. It follows from these predictions that populations with stronger sexual selection would have fitness benefits, allowing them to adapt to novel environments [*reviewed in* @Candolin_2008]. Additionally, sexual selection can reinforce local adaptations through mate choice on phenotypes that effectively advertise quality in a specific ecological context [@Boughman_2002; @Maan_2011]. For instance, beak morphology is a trait that arose under natural selection in several taxa and is now a trait under sexual selection [a.k.a. *magic traits*; @Servedio_2011]. In Darwin's finches (*Geospiza fortis*) divergent beak morphology is an adaptation to local food availability that has been maintained through assortative mating [@Huber_2007]. Through these synergetic effects sexual selection and natural selection (*e.g.* environmental change over space and time) may lead to increased adaptation and speciation rates.

As a speciose and well-characterised group of organisms with remarkable and variable sexually-selected traits such as song and colourful plumage, birds are a popular focus for macroevolutionary studies of sexual selection and diversification [*e.g.*, @Barraclough_1995; @Seddon_2008; @Huang_2014; @Seddon_2013; @Morrow_Speciation_2003]. A 2011 meta-analysis, covering 20 studies of birds and other taxa, found a small but significant positive association between sexual selection and speciation, with the average effect size in birds stronger than in mammals but weaker than in insects and fish [@Kraaijeveld_2011]. However, @Kraaijeveld_2011 found large variation in effect sizes between studies, likely because of differences in the metrics used as estimates of speciation and the strength of sexual selection. More recently, @Huang_2014 found no association between sexual dichromatism and speciation (*n* = 918 species) when using spectrophotometry measures on taxonomic specimens [@Armenta_2008] and tip-rate estimates from a molecular-only phylogeny [@Jetz_2012]. Similarly, @Cooney_2017 found no effect of sexual dichromatism on diversification across 1,306 pairs of species, using dichromatism scores from human observers.

Here, we investigate the association between sexual selection and diversification in birds while extending previous work in multiple ways. Firstly, many macroevolutionary studies estimate the strength of sexual selection using proxies such as sexual dimorphism or dichromatism, but these are imperfect correlates. We used a very large (*n* = 5,812) dataset of sexual dichromatism [estimated from illustrated drawings; @Dale_2015], as well as a smaller but still substantial dataset (*n* = 2,465) that gives a composite measure of male-biased sexual selection, capturing variation in sexual dimorphism, social polygyny and [lack of] paternal care [@Dale_2015]. These datasets allow us to cover the great majority of bird genera, in contrast to previous work that has often focused on particular clades [*e.g.* @Seddon_2008]. Secondly, we use the latest methods for estimating speciation and extinction rates at the tips of phylogenetic trees, including BAMM [Bayesian Analysis of Macroevolutionary Mixtures; *see*, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017], as well as older but less controversial tip-rate statistics, such as diversification rate ($\lambda_{DR}$) and node density ($\lambda_{ND}$). We also conduct analyses that allow phylogenetic uncertainty in the avian tree [@Jetz_2012; *but see* @Rubolini_2015] to affect the precision of our estimates, rather than implicitly assuming that the single best-supported phylogenetic tree, or consensus tree, is correct. Thirdly, our analysis includes several potentially important environmental variables as predictors: range size, seasonal variation in temperature, spatial temperature variation (across a species’ range), long-term variation in temperature between now and the last interglacial (LIG) as well net primary productivity (NPP). By including these environmental variables in our models, we statistically account for their effects on variation in diversification rates and thereby improve our estimates of the effects of sexual selection. We also test whether the effects of sexual selection are contingent on the environment, as predicted by some theories [*e.g.* the theory that sexual selection is more likely to elevate adaptation in changeable relative to static environments; @Long_2012; @Connallon_Hall_2016].  \phantomsection

#Materials and methods

@@

##Compiling data for sexual selection and environmental stress
\phantomsection

###Sexual dichromatism

We used a previously published measure of sexual dichromatism for 5,983 species of passerines [@Dale_2015]. Briefly, @Dale_2015 obtained sex-specific RGB (red, green and blue) values across six body patches (nape, crown, forehead, throat, upper breast, and lower breast) from _Handbook of the Birds of the World_ [@delhoyo2011]. The relative contribution of male and female RGB colour values were averaged across body patches and provide 'male-like' and 'female-like' plumage scores. Here we use the absolute difference between male and female plumage scores as an estimate of sexual dichromatism. Additionally, we used a more accurate measure of dichromatism corresponding to colour distance in avian colour space derived from spectral data [@Armenta_2008]. These measurements include variation in the ultraviolet and bird-visible range and --- unlike the RGB measures --- are sourced from taxonomic samples (as opposed to colour drawings). However, this spectrophotometry data is not as extensive as the RGB measures; with only 581 passerine species available for this analysis. While there is a correlation between these two measures, there is residual variation ($r = 0.79$; \href{https://justincally.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S10}).  \phantomsection

###Male-biased sexual selection

Sexual dichromatism is a widely used indicator of sexual selection in birds [*e.g.*, @Barraclough_1995; @Seddon_2013; @Morrow_Speciation_2003; @Owens_1999; @Huang_2014], but only partially captures variation in the strength of sexual selection across taxa. Therefore, we also use an existing dataset [@Dale_2015] of male-biased sexual selection based on the first component of a phylogenetic principle component analysis (PPCA) of three characteristics associated with sexual selection (sexual size dimorphism, social polygyny and [lack of] paternal care). This measure of male-biased sexual selection is available for only 2,465 species and is weakly correlated with the absolute values of sexual dichromatism using RGB measures ($r = 0.34$; \href{https://justincally.github.io/SexualSelection_Speciation/#analysis_using_male-biased_measure_of_sexual_selection}{Figure S12}).  \phantomsection

###Environmental variables

@@

###Generating biologically relevant predictors for environmental variation

Using the extracted environmental variables from each species range size we developed biologically meaningful predictors of environmental variation/stress relating to (*i*) seasonal climate variation, (*ii*) spatial climate variation and (*iii*) long-term climate variation. To obtain seasonal climate variation we used mean values of temperature seasonality (BIO4) for each range. To estimate levels of spatial environmental variation a species may endure we used the first principle component (PC1) from a PCA on standard deviations from all bioclimatic variables, excluding temperature (BIO4) and precipitation seasonality (BIO15). PC1 was heavily loaded towards bioclimatic variables relating to temperature, thus PC1 reflects the variation in temperature across a species' range (\href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Table S1}). Given that species range is a potentially informative predictor of speciation and extinction, we controlled for the correlation between spatial variation and range size --- where larger ranges have larger variation in PC1 --- by taking the residuals of a fitted general additive model (GAM; \href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Figure S1}). To obtain long-term variation in climates for each species range we take the first principal component of the absolute difference in the bioclimatic variables between the LIG and current values. Similarly to spatial variation, the long-term climate variation is primarily loaded to temperature differences between the LIG and current climates (\href{https://justincally.github.io/SexualSelection_Speciation/#long_term_climate_variability_(lig_anomaly)}{Table S2, Figure S2}). The five predictors of environmental variability are not strongly correlated (\href{https://justincally.github.io/SexualSelection_Speciation/#correlations_between_environmental_predictors}{Figure S3}). Details and ``R`` code to generate these predictors can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM}.  \phantomsection

##Estimating extinction and speciation

Phylogenetic information was obtained from www.birdtree.org [@Jetz_2012]. We used a maximum clade credibility (MCC) tree from 2,500 samples of the posterior distribution that were subset to the passerine bird order (*n* = 5,965) as the main phylogenetic tree in our comparative analysis. Additionally, a random draw of 100 full trees (including species without genetic data) from the posterior distribution of phylogenetic trees was used for diversification analyses using tip-rate measures and BAMM [@Rabosky_2014]. These trees used a 'Hackett backbone' [@Hackett_2008] and were constructed using a pure birth (Yule) model. We calculated three different tip-rate metrics of speciation and one of extinction across all trees.

Firstly, we obtained two tip-rate metrics of speciation using statistics derived from the properties of the nodes and branches along root-to-tip paths of the phylogeny. Node density (ND) is a simple statistic calculating the density of nodes from the phylogenetic root to the tip, while the log-transformed equal splits (logES), also known as diversification rate (DR) [*e.g.*, @Jetz_2012; @quintero_2018; @Rabosky_fish_2018], is derived from the sum of edge lengths branching from a node, with each more basal node having the sum of lengths down-weighted. Studies have suggested that DR and ND (henceforth referred to as $\lambda_{DR}$ and $\lambda_{ND}$) are more reflective estimates of speciation than diversification [@Belmaker_2015; @Rabosky_diversification_2018]. Therefore, $\lambda_{DR}$ is a measure of speciation rate more heavily weighted to recent speciation events and $\lambda_{ND}$ is a measure of speciation across the root-to-tip path. These tip-rate measures are alternatives to state-dependent diversification models such as Quantitative State Speciation-Extinction (QuaSSE). But, based on previous simulation studies, $\lambda_{DR}$ and $\lambda_{ND}$ are robust and intuitive measures that provide high power and low false discovery rate with large phylogenies when incorporated into Phylogenetic Generalized Least Squares (PGLS) models [@Harvey_essim_2017].

Secondly, we used BAMM to model the dynamics of speciation and extinction across the 100 phylogenetic trees [@Rabosky_2014]. This software uses a Bayesian approach (reversible-jump Markov Chain Monte Carlo) to generate probability distributions of evolutionary rate-shift configurations with variable speciation and extinction rates. Importantly, these models provide tip-rate estimates of speciation and extinction rate. The parameters of the 100 BAMM runs are detailed in full in the \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM}; briefly, we used a time-variable model with 100 expected number of shifts and prior rates set from the initial speciation and extinction values using the ``BAMMtools`` R package [@bammtools_2014]. Each BAMM model (one MCC and 100 random draws of the posterior) was run for 100 million generations, and given the computationally intensive nature of BAMM, runs were conducted across multiple CPUs. Each run of BAMM reached convergence with effective sample size (ESS) of MCMC (Markov Chain Monte Carlo) samples surpassing 200 (an arbitrary value, above which posterior distributions can often be accurately inferred) for the key model parameters of log-likelihood and number of rate shifts (\href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S3}, \href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S4}). Further details of BAMM parameters and output are available in the \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM}, with tip-rate means and variances provided. Additionally, given the variability in BAMM estimates, we also provide analysis of BAMM shift configurations and tip-rate estimates from our run on the MCC tree and within a BAMM run on the MCC tree from a genetic-only phylogeny across all birds [@Harvey_2017]. All analyses were conducted on log-rates.   \phantomsection

##Phylogenetic comparative analysis

>>>>

---
title: \huge{Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds}
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§}
  and Iliana Medina Guzmán\textsuperscript{§}
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang,
  flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{colorlinks=true, hyperfootnotes=false,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: Sexual selection can shape phylogenetic diversity by affecting speciation or extinction rates. The predicted effects on diversity are often contradictory; for example, sexual selection might promote speciation by creating reproductive isolation, or hinder it by selecting for traits that facilitate gene flow. Here, we investigate the relationship between sexual selection and diversification in passerine birds. Multiple studies use sexual dichromatism as a proxy for sexual selection, however,  this relationship can be weak in many clades. Thus, we use two measures of sexual selection --- sexual dichromatism and a multivariate measure of male-biased sexual selection --- to test the link between sexual selection and diversification. We also test whether the effect of sexual selection on diversification is contingent on ecological variables. Our results show that male-biased sexual selection, but not sexual dichromatism, can explain speciation rates ($\lambda_{DR}$) in passerines. We also find that birds with smaller ranges have higher speciation rates, but there is no interaction between environmental variables and sexual selection. Our findings show that sexual dichromatism is a poor proxy for sexual selection at least a a broad-scale, and support the view that sexual selection is an evolutionary force shaping patterns of diversity amongst songbirds. \vspace{5mm}  \par\noindent
  \textbf{Keywords:} Speciation rate, Comparative analysis,Plumage dichromatism, Range size, Extinction rate.
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

# Introduction

Sexual selection results from competition for matings, fertilisations, or associated resources, and is typically stronger in males than females [@Fromhage_2016; @Janicke_2016; @Kokko_Jennions2008]. This evolutionary force has been proposed to facilitate speciation through the evolution of divergent signals associated with improved mating success [@Lande_1981; @Lande_1982]. Experiments suggest that sexual selection can influence the evolution of a surprisingly diverse set of traits [@Cally_2019], and may therefore strongly influence extant patterns of species diversity. For instance, in antbirds (Thamnophilidae), genera with complex songs and striking dichromatism are more species-rich [@Seddon_2008]. Divergent mating signals in one sex (usually males) co-evolves with divergent preferences for those signals in the opposite sex, leading to behavioural reproductive isolation [@Safran_2013].

Alongside facilitating the evolution of divergent mating signals, sexual selection can increase the extent populations diverge by enlarging the available phenotypic space for mating signals. These predictions emerge as sexual selection may increase genetic diversity as a result of trade-offs between naturally- and sexually-selected traits or trade-offs under *intra*-locus sexual conflict between male and female trait expressions [@Lorch_2003; @Bonduriansky_2011; @Radwan_2016]. Additionally, sexual selection can have both positive and negative consequences for adaptation [*e.g.*, @Kokko_Brooks_2003; @Whitlock_Agrawal_2009; @Holman_Kokko_2013]. For example, sexual selection can elevate average fitness by helping to purge deleterious mutations [@Agrawal_2001; @Siller_2001] and fix beneficial ones [@Whitlock_2000], which should mitigate extinction risk and facilitate adaptation to environmental change. However, sexual selection might also promote extinction by selecting for male traits that have detrimental pleiotropic effects on female traits [@Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013; @Pischedda_2006], or selecting for male phenotypes with improved mating success but which harm population productivity, such as under-investing in parental care [@RN277].

Importantly, the effects of sexual selection on adaptation and speciation may depend on environmental conditions. Both theoretical and empirical studies suggest that under stressful/changing environments, sexual selection may have greater fitness benefits than under benign environments. Stressful environments strengthen the positive correlation between male mating success and female fitness, reducing the burden of negative pleiotropic effects [@Berger_2014; @Connallon_Hall_2016; @Long_2012; @Martinossi-Allibert_Savković_2017]. Conversely, benign or stable environments ensure consistent selection, preferentially eroding genetic variation at sexually concordant loci (*i.e.* loci where the optimum genotype is equivalent for both sexes). In these stable environments, genetic variation remains disproportionately at sexually antagonistic loci; causing displacement of male and female traits from their optimum. It follows from these predictions that populations with stronger sexual selection would have fitness benefits, allowing them to adapt to novel environments [*reviewed in* @Candolin_2008]. Additionally, sexual selection can reinforce local adaptations through mate choice on phenotypes that effectively advertise quality in a specific ecological context [@Boughman_2002; @Maan_2011]. For instance, beak morphology is a trait that arose under natural selection in several taxa and is now a trait under sexual selection [a.k.a. *magic traits*; @Servedio_2011]. In Darwin's finches (*Geospiza fortis*) divergent beak morphology is an adaptation to local food availability that has been maintained through assortative mating [@Huber_2007]. Through these synergetic effects sexual selection and natural selection (*e.g.* environmental variation over space and time) may lead to increased speciation rates.

Birds are a speciose and well-characterised group of organisms with remarkable and variable sexually-selected traits such as song and colourful plumage, thus are a popular focus for macroevolutionary studies of sexual selection and diversification [*e.g.*, @Barraclough_1995; @Seddon_2008; @Huang_2014; @Seddon_2013; @Morrow_Speciation_2003]. A 2011 meta-analysis, covering 20 studies of birds and other taxa, found a small but significant positive association between sexual selection and speciation, with the average effect size in birds stronger than in mammals but weaker than in insects and fish [@Kraaijeveld_2011]. However, @Kraaijeveld_2011 found large variation in effect sizes between studies, likely because of differences in the metrics used as estimates of speciation and the strength of sexual selection. More recently, @Huang_2014 found no association between sexual dichromatism and speciation (*n* = 918 species) when using spectrophotometry measures on taxonomic specimens [@Armenta_2008] and tip-rate estimates from a molecular-only phylogeny [@Jetz_2012]. Similarly, @Cooney_2017 found no effect of sexual dichromatism on diversification across 1,306 pairs of species, using dichromatism scores from human observers.

Here, we investigate the association between sexual selection and diversification in birds while extending previous work in multiple ways. Firstly, multiple macroevolutionary studies estimate the strength of sexual selection using proxies such as sexual dimorphism or dichromatism, but sexual selection in birds can involve many other traits, such as song, or the use of ornaments like long tail feathers, bowers, crests and displays [*e.g.*, @Uy_2000; @Pryke_2001]. In fact, it was recently highlighted that song divergence is inversely correlated to sexual dichromatism in passerines, suggesting that there is a trade-off between signalling modalities [@Cooney_2018]. Our study compares the role of sexual dichromatism and other sexual selection proxies in the generation of species. Additionally, we test the relative contribution of environmental variables to the generation of diversity, and how environment and sexual selection could interact to drive speciation processes [@Long_2012; @Connallon_Hall_2016].

To test the link between different measures of sexual selection and diversification processes we use a large (*n* = 5,812) dataset of sexual dichromatism [estimated from illustrated drawings; @Dale_2015], as well as a reduced but still substantial dataset (*n* = 2,465) that gives a composite measure of male-biased sexual selection, capturing variation in sexual dimorphism, social polygyny and [lack of] paternal care [@Dale_2015]. These datasets allow us to cover the majority of bird genera. Additionally, we use multiple metrics for estimating speciation and extinction rates at the tips of phylogenetic trees, including BAMM [Bayesian Analysis of Macroevolutionary Mixtures; *see*, @Beaulieu_2015; @Rabosky_Extinction_2016; @Moore_2016; @Rabosky_BAMM_flawed_2017], as well as older but reliable tip-rate statistics, such as diversification rate ($\lambda_{DR}$) and node density ($\lambda_{ND}$).  \phantomsection

# Materials and methods

@@

## Compiling data for sexual selection and environmental stress
\phantomsection

### Sexual dichromatism

We used a previously published measure of sexual dichromatism for 5,983 species of passerines [@Dale_2015]. Briefly, @Dale_2015 obtained sex-specific RGB (red, green and blue) values across six body patches (nape, crown, forehead, throat, upper breast, and lower breast) from _Handbook of the Birds of the World_ [@delhoyo2011]. The relative contribution of male and female RGB colour values were averaged across body patches and provide 'male-like' and 'female-like' plumage scores. Here we use the absolute difference between male and female plumage scores as an estimate of sexual dichromatism. Additionally, we used another measure of dichromatism corresponding to colour distance in avian colour space derived from spectral data [@Armenta_2008]. These measurements include variation in the ultraviolet and bird-visible range and --- unlike the RGB measures --- are sourced from museum samples (as opposed to colour drawings). However, this spectrophotometry data is not as extensive as the RGB measures; with only 581 passerine species available for this analysis. While there is a correlation between these two measures, there is residual variation ($r = 0.79$; \href{https://justincally.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S10}).  \phantomsection

### Male-biased sexual selection

Sexual dichromatism is a widely used indicator of sexual selection in birds [*e.g.*, @Barraclough_1995; @Seddon_2013; @Morrow_Speciation_2003; @Owens_1999; @Huang_2014], but only partially captures variation in the strength of sexual selection across taxa. Therefore, we also use an existing dataset [@Dale_2015] of male-biased sexual selection based on the first component of a phylogenetic principle component analysis (PPCA) of three characteristics possitively associated with sexual selection (sexual size dimorphism, social polygyny and [lack of] paternal care). This measure of male-biased sexual selection is available for only 2,465 species and is weakly correlated with the absolute values of sexual dichromatism using RGB measures ($r = 0.34$; \href{https://justincally.github.io/SexualSelection_Speciation/#analysis_using_male-biased_measure_of_sexual_selection}{Figure S12}).  \phantomsection

### Environmental variables

@@

### Generating biologically relevant predictors for environmental variation

Given that stressful environments are expected to interact with sexual selection and have a positive effect on adaptation [@Cally_2019], we used the extracted environmental variables from each species range size to develop biologically meaningful predictors of environmental variation/stress. Firstly we used (*i*) the average NPP in each species' range and (*v*) the log-transformed range size as potentially informative predictors of speciation rates. We also used three environmental predictors dervived from bioclimatic data. These predictors relate to seasonal climate variation, spatial climate variation and long-term climate variation. To obtain seasonal climate variation we used (*iii*) mean values of temperature seasonality (BIO4) for each range. (*iV*) To estimate levels of spatial environmental variation a species may endure we used the first principle component (PC1) from a PCA on standard deviations from all bioclimatic variables, excluding temperature and precipitation seasonality (BIO 4 and BIO15). PC1 was heavily loaded towards bioclimatic variables relating to temperature, thus PC1 reflects the variation in temperature across a species' range (\href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Table S1}). Species range is a potentially informative predictor of speciation and extinction, so we controlled for the correlation between environmental spatial variation and range size --- where larger ranges have larger variation in PC1 --- by using the residuals of a fitted general additive model (GAM; \href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Figure S1}) as a predictor. To obtain long-term variation in climates for each species range we take (*v*) the first principal component of the absolute difference in the bioclimatic variables between the LIG and current values. Similarly to spatial variation, the long-term climate variation is primarily loaded to temperature differences between the LIG and current climates (\href{https://justincally.github.io/SexualSelection_Speciation/#long_term_climate_variability_(lig_anomaly)}{Table S2, Figure S2}). The five predictors of environmental variability are not strongly correlated (\href{https://justincally.github.io/SexualSelection_Speciation/#correlations_between_environmental_predictors}{Figure S3}). Details and ``R`` code to generate these predictors can be found within the \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM}.  \phantomsection

## Estimating extinction and speciation

Phylogenetic information was obtained from www.birdtree.org [@Jetz_2012]. We used a maximum clade credibility (MCC) tree from 2,500 samples of the posterior distribution (*n* = 5,965) as the main phylogenetic hypothesis in our comparative analysis. Additionally, a random draw of full trees (including species without genetic data) from the posterior distribution of phylogenetic trees was used for diversification analyses using tip-rate measures (1000 trees) and BAMM (100 trees)[@Rabosky_2014]. These trees used a 'Hackett backbone' [@Hackett_2008] and were constructed using a pure birth (Yule) model. We calculated three different tip-rate metrics of speciation and one of extinction across all trees.

Firstly, we obtained two tip-rate metrics of speciation using statistics derived from the properties of the nodes and branches along root-to-tip paths of the phylogeny. Node density (ND) is a simple statistic calculating the density of nodes from the phylogenetic root to the tip, while the log-transformed equal splits (logES), also known as diversification rate (DR) [*e.g.*, @Jetz_2012; @quintero_2018; @Rabosky_fish_2018], is derived from the sum of edge lengths branching from a node, with each edge towards the root having the length down-weighted. Studies have suggested that DR and ND (henceforth referred to as $\lambda_{DR}$ and $\lambda_{ND}$) are more reflective estimates of speciation than diversification [@Belmaker_2015; @Rabosky_diversification_2018]. Therefore, $\lambda_{DR}$ is a measure of speciation rate more heavily weighted to recent speciation events and $\lambda_{ND}$ is a measure of speciation across the root-to-tip path. These tip-rate measures are alternatives to state-dependent diversification models such as Quantitative State Speciation-Extinction (QuaSSE). But, based on previous simulation studies, $\lambda_{DR}$ and $\lambda_{ND}$ are robust and intuitive measures that provide high power and low false discovery rate with large phylogenies when incorporated into Phylogenetic Generalized Least Squares (PGLS) models [@Harvey_essim_2017].

Secondly, we used BAMM to model the dynamics of speciation and extinction across the 100 phylogenetic trees [@Rabosky_2014]. This software uses a Bayesian approach (reversible-jump Markov Chain Monte Carlo) to generate probability distributions of evolutionary rate-shift configurations with variable speciation and extinction rates. Importantly, these models provide tip-rate estimates of speciation and extinction rate. The parameters of the 100 BAMM runs are detailed in full in the \href{https://justincally.github.io/SexualSelection_Speciation/#set_up_bamm_parameters}{ESM}; briefly, we used a time-variable model with 100 expected number of shifts and prior rates set from the initial speciation and extinction values using the ``BAMMtools`` R package [@bammtools_2014]. Each BAMM model (one MCC tree and 100 random draws of the posterior) was run for 100 million generations, and given the computationally intensive nature of BAMM, runs were conducted across multiple CPUs. Each run of BAMM reached convergence with effective sample size (ESS) of MCMC (Markov Chain Monte Carlo) samples surpassing 200 (an arbitrary value, above which posterior distributions can often be accurately inferred) for the key model parameters of log-likelihood and number of rate shifts (\href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S3}, \href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S4}). Further details of BAMM parameters and output are available in the \href{https://justincally.github.io/SexualSelection_Speciation/}{ESM}, with tip-rate means and variances provided. Additionally, given the variability in BAMM estimates, we also provide analysis of BAMM shift configurations and tip-rate estimates from our run on the MCC tree and within a BAMM run on the MCC tree from a genetic-only phylogeny across all birds [@Harvey_2017]. All analyses were conducted on log-rates.  \phantomsection

## Phylogenetic comparative analysis
